### 2017-06-30 motivation.tex
Motivating automation.

The second observation that informs our approach is one that pertains
to automation. Program verification, even when machine-aided, often
entails significant annotation burden in the form of intermediary
assertions and loop invariants required to prove a program correct.
Even with concurrent program logics, such as Rely-Gurantee, which
extend Hoare logic with additional artifacts, (stable) intermediary
assertions and loop invariants remain a major source of annotation
burden. Since database transactions inherently deal with the mutable
state of the database, it is natural to formulate them as concurrent
imperative programs. An advantage with this formulation is that it
lets us immediately apply existing concurrent program logics to
database transactions, albeit with necessary customizations to deal
with weak isolation. The downside, however, is that there hardly is
any automation support to apply such program logics on concurrent
imperative programs. Attemps at automation face the problem of
inferring stable and inductive loop invariants in its full generality,
which, while being undecidable in theory, and hard in practice, also
takes our focus away from weak isolation.

In the light of aforementioned reservations, we refrain from
formulating database transactions as concurrent imperative programs.
Instead, we see value in viewing them as essentially functional
programs that manage the database state via a monad akin to the
\C{State} monad~\cite{statemonad}. We find it useful to reason about
statements that mutate the database state, not in terms of a pre- and
post-condition pair, but in terms of a state transformer that relates
the pre- and post-states of a statement. Loops are substituted with
higher-order combinators that have a well-defined control flow, and
facilitate well-founded recursion. The static semantics of these
combinators let us lift the state transformer of its higher-order
argument (i.e., the loop body) to the state transformer of the
combined expression (i.e., the loop). This is made possible due to the
characteristic structure of database programs, which also gets
reflected in the structure of the state transformer. We illustrate
this point through a simple example that doesn't involve transactions,
but is nonetheless relevant.

% \begin{figure}[!t]
% \centering
% %
% \begin{subfigure}[b]{0.46\textwidth}
% \begin{ocaml}
% Set s' = $\emptyset$;
% foreach x in s {
%   s'.add(f(x)); 
% }
% \end{ocaml}
% \caption{}
% \end{subfigure}
% %
% \begin{subfigure}[b]{0.5\textwidth}
% \begin{ocaml}
% let s' = ref [];
% foreach s (fun x -> s' := x::(!s'));
% \end{ocaml}
% \caption{Lorem ipsum, lorem ipsum,Lorem ipsum, lorem ipsum,Lorem ipsum}
% \end{subfigure}
% \caption{Caption place holder}
% %
% \caption{New versions are created from existing versions either
% through \C{push} or \C{merge}.}
% \label{fig:syntactic-ancestors}
% \end{figure}
\begin{figure}[!h]
\begin{ocaml}
let l' = ref [];
foreach l (fun x -> l' := x::(!l'));
\end{ocaml}
\caption{A program that uses mutable state to reverse a list of elements}
\label{fig:mutable-rev}
\end{figure}

Consider the program in Fig.~\ref{fig:mutable-rev} that populates a
mutable list (\C{l'}) with elements from other list (\C{l}), reversing
their order. The aim is to show that, when the program ends, both
\C{l'} and \C{l} contain the same set of elements. If we were to
attempt Hoare-style verification, we are obligated to define an
inductive invariant for the \C{foreach} loop. Unfortunately, none
exists for the loop, as such. The reason is that there is no way in
the program to refer to the elements in the list that have already
been added to the list \C{l'}. This problem can be easily rectified by
introducing an integer \C{i} that keeps track of the index of \C{x} in
\C{l}.

>>>>

\input{motiv-automation}
